The Private School Network: Recruiting Visits to Private High Schools by Public and Private Universities

### 2024-02-16 index.qmd
working through presentation

# Motivation

### Motivation

#### Off-campus recruiting visits to high schools in Nashville

@@

### Motivation

-   Recruiting visit indicates university and school have social relationship
-   Visits (edges) from university (mode 1) to high school (mode 2) yields a two-mode social network
-   Comparing recruiting network of public research and selective private universities gives insight about competition for affluent students

@@

### Research on nonresident enrollment at public universities

@@

- Institutional financial aid [@RN3762]
- Recruiting to solicit 'inquiries' and applications
  - "Student list" purchases [@RN4985]
  - Off-campus recruiting visits [@RN4758; @RN4934]
    - Out-of state visits to public high schools focus on affluent public schools, avoid non-white communities

### Research on links between private schools and private universities

Studies find attending private school increases college access, selectivity

- Public school students: 5% attend "top 100" public univ; 2% attend top 100 private univ (Author calculations from HSLS 2009)
- Private school: 11% attend top 100 public; 17% attend top 100 private
- Nonsectarian private: 11% attend top 100 public; 40% attend top 100 private

Research from economics [@RN4935]

- Rich students have higher non-academic ratings because they attend elite private schools (extracurricular activities, recommendations)

Research from sociology highlights organizational linkages [@RN4935]

### slide title
#### slide sub-title

@@

## Data and Sample

@@

## Constructing Social Networks

## Analyses

>>>>

# Introduction

### Off-campus recruiting visits to high schools in Nashville

@@

### Setup

Recruiting visit indicates university and school have social relationship

- The university wants to enroll some students from the school
- The school wants to send some students to the university

Visits from universities to schools yields a two-mode social network

- "Edges" = visits; "Vertices" = schools (mode 1) and universities (mode 2)

@@

### Literature Review
#### Nonresident enrollment at public universities

@@

- Institutional financial aid to convert admits [@RN3762]
- Recruiting interventions at earlier stages of enrollment funnel
  - "Student list" purchases [@RN4985]
  - Off-campus recruiting visits [@RN4758; @RN4934]
    - Out-of state visits to public high schools focus on affluent public schools, avoid non-white communities

### Literature Review
#### Linkages between private schools and private universities
<hr>
Studies find attending private school increases college access, selectivity

- Public school students: 5% attend "top 100" public univ; 2% attend top 100 private univ (Author calculations from HSLS 2009)
- Private school: 11% `@` top 100 public; 17% `@` top 100 private
- Nonsectarian private: 11% `@` top 100 public; 40% `@` top 100 private

@RN4935 study Ivy Plus admissions decisions

- Rich students have higher non-academic ratings because they attend elite private schools (extracurriculars, recommendations)

Sociology highlights school-college organizational linkages

- Visits maintain relations with counselors at feeder schools [@RN3519]
- Private school counselors lobby for marginal candidates [@RN4407]

### Literature Review
#### Synthesis
<hr>
These two literatures have developed in isolation

- Pursuing out-of-state private school students by public research univs encroaches on historic territory of selective private univs

Analyzing network of recruiting visits to private schools gives insight about competition for affluent students

- Public and private universities may be targeting similar sets of schools
  - Direct competition or vying to be second choice for rejected applicant

Private school students more likely to be affluent, white than public school students [@RN4504]

- Competitive recruiting of private school students has downstream effects on racial and socioeconomic composition of colleges

@@

### Methods
#### Data and Sample

Data collection (see @RN4758)

- 2017 off-campus recruiting visits scraped from univ admissions websites

Data collection sample

- Public "research-extensive" universities from 2000 Carnegie (N=102)
- Private universities in top 100 of U.S. News *National Universities* (N=57)
- Private colleges in top 50 of *National Liberal Arts Colleges* (N=47)

Analysis sample: univs that posted complete recruiting visits on website

- 15 public research, 14 private universities, 12 private liberal arts (omitted)

Secondary data

- IPEDS; CCD; Private School Survey; Niche *Best Private High Schools*

@@

### Methods
#### Constructing social networks

text
- Transform two-mode network to one-mode university network showing number of schools visited by both university $i$ and university $j$

### Methods
#### Analyses by research question

1.  How does the scale of visits to private high schools vary for the public research universities and selective private universities?
    - Simple descriptive statistics

\medskip

2.  To what extent do public research universities and selective private universities visit overlapping sets of private high schools?
    - Graph one-mode university network; universities allocated to categories based on cluster analysis

\medskip

3.  What are the characteristics of private schools that receive visits from public research universities and selective private research universities?
    - Descriptive statistics of characteristics of university order-1 "ego network" (schools the university visited):
      - Geographic region, religion, racial composition, academic reputation

### 2024-02-16 r_r/submission/manuscript_rihe_r_r.Rmd
working through presentation

Although the literature on nonresident enrollment developed in isolation from the literature on private school to private university linkages, both phenomena are part of a broader story about the changing nature of school segregation and the competition for affluent, traditional college students. Drawing from resource dependence theory, as public research universities become more reliant, they have a financial incentive to target out-of-state private schools, which tend to enroll affluent students that can pay high tuition prices. Although a handful of public research universities have relied on nonresident enrollment for many decades (e.g., University of Michigan, University of Delaware), public research universities began substantially increasing nonresident enrollment en masse around 2005 [@RN3753]. By contrast, selective private universities have long relied on enrollment from private schools [@RN3519; @RN4640]. Therefore, the pursuit of out-of-state private school students by public research universities is a relatively recent encroachment on the historic territory of selective private universities and in-state flagships. Speaking about this competition, an enrollment management consultant said, "'Everybody wants to go to the magic island of full pay students, but it’s rapidly shrinking real estate'" [@RN4911]. By analyzing the network of recruiting visits, we show whether public research universities and selective private universities are competing for students at similar sets of private schools. This may be direct competition for the same student or vying to be the second choice for students who are rejected by their top choice.

>>>>

Although the literature on nonresident enrollment developed in isolation from the literature on private school to private university linkages, both phenomena are part of a broader story about the changing nature of school segregation and the competition for affluent, traditional college students. Drawing from resource dependence theory, as public research universities become more tuition reliant, they have a financial incentive to target out-of-state private schools, which tend to enroll affluent students that can pay high tuition prices. Although a handful of public research universities have relied on nonresident enrollment for many decades (e.g., University of Michigan, University of Delaware), public research universities began substantially increasing nonresident enrollment en masse around 2005 [@RN3753]. By contrast, selective private universities have long relied on enrollment from private schools [@RN3519; @RN4640]. Therefore, the pursuit of out-of-state private school students by public research universities is a relatively recent encroachment on the historic territory of selective private universities and in-state flagships. Speaking about this competition, an enrollment management consultant said, "'Everybody wants to go to the magic island of full pay students, but it’s rapidly shrinking real estate'" [@RN4911]. By analyzing the network of recruiting visits, we show whether public research universities and selective private universities are competing for students at similar sets of private schools. This may be direct competition for the same student or vying to be the second choice for students who are rejected by their top choice.
